### 2013-06-11 context-dependence.tex
writeup

We clearly don't want to add an extra parameter $\pi(z)$ for all $z \in \calS^{m+2\ell}$. 
If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
One approach would be to put a Dirichlet distribution on $\pi(z)$,
but this would destroy many of the nice computational properties of the likelihoods,
since making the frequencies of patterns at the root random introduces correlations between counts.
An alternative is to specify $\pi$ through lower-dimensional statistics,
for instance to model the root state as a sample from a Markov random field
with parameters for the marginal frequencies of each state, and pairwise (or higher) correlations between them.

@@

@@

%%%
\subsection{CpG mutation}

Nucleotide sequences of length $10^6$ were simulated with various

@@

  \item Do inference along a tree.

@@

  \item See what's feasible with a larger alphabet -- 4 bases, or 20 amino acids -- and more parameters.

  \item Show that taking $\ell=0$ can be positively misleading (when?).

>>>>

We clearly don't want to add an extra parameter $\pi(z)$ for all $z \in \calS^{m+2\ell}$. 
If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
One approach would be to put a Dirichlet distribution on $\pi(z)$.
This has the nice property that you can choose priors that are consistent across different values of $m$ and $\ell$.
We'd have to still MCMC around in the space of $\pi(z)$, thought, so they'd be nuisance parameters from that point of view.
Also, making the frequencies at the root random introduces correlations between counts, which we want to avoid.
An alternative is to specify $\pi$ through lower-dimensional statistics,
for instance to model the root state as a sample from a Markov random field
with parameters for the marginal frequencies of each state, and pairwise (or higher) correlations between them.

@@

\begin{figure}
  \begin{center}
    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
  \end{center}
  \caption{ 
  MCMC traces of the parameters in the Ising model across 100,000 iterations.
  The units of $\lambda$ are in mean number of changes per site.
  Vertical line shows the true value.
  }
\end{figure}

@@

%%%
\subsection{CpG mutation, on a tree}

Nucleotide sequences of length $10^6$ were simulated down two (unequal) branches of a tree,
with all single-base substitution rates $m_{xy}$ set equal (varying from a total of .01 to .1 across the tree)
and an additional CpG rate $\gamma$ set to either 0 or three times the single-base rate.
Analysis was done with $\ell=2$ and $w=1$ (a five-to-one window).

The state at the root of the tree was independently chosen bases with equal frequencies, 
and these frequencies at the root were added as parameters to the model (with a Dirichlet prior).

The point estimates are good, 
as are the MCMC estimates,
but there seems to be mixing trouble having to do with the root frequencies.

\begin{figure}
  \begin{center}
    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-initfreqs}
  \end{center}
  \caption{ 
  MCMC traces across 660,000 iterations of the frequencies at the root of the (two-taxon) tree in the CpG model.
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics{writeup-plots/selsims-2013-06-03-13-17-0790276-mutrates}
  \end{center}
  \caption{ 
  MCMC traces across 660,000 iterations of several mutation rate parameters in the CpG model.
  Units are in mean number of substitutions per base across the entire tree.
  Vertical line shows the true value.
  }
\end{figure}

@@

  \item Allow more than single-base distribution at the root.

  \item What's up with the mixing problem in root frequencies?

@@

  \item See what's feasible with a larger alphabet -- 60 codons -- and more parameters.

  \item Show that taking no context ($\ell=0$) can be positively misleading in some context.

  \item Do a three-taxa tree.

  \item How to identify motifs in the residuals?
